Two group comparisons and data aggregation
======================================================================


```{r include = FALSE}
library(knitr)
opts_chunk$set(tidy = FALSE)
```


```{r}
library(lattice)
library(ggplot2)
library(plyr)
library(reshape)
prDat <- read.table("../data/photoRec/GSE4051_data.tsv")
str(prDat, max.level = 0)
prDes <- readRDS("../data/photoRec/GSE4051_design.rds")
str(prDes)
```

## Two sample tests -- one gene

Let's extract the data for one gene and put in a data.frame with the experimental information.
```{r}
set.seed(987)
(theGene <- sample(1:nrow(prDat), 1))
pDat <- data.frame(prDes, gExp = unlist(prDat[theGene, ]))
str(pDat)
```

Always explore the data before plunging into analysis! What are the sample means in the wild type and Nrl knockout groups (yes, we're ignoring developmental stage today)? (`aggregate()` and other data aggregation functions are explained below.)

```{r}
aggregate(gExp ~ gType, pDat, FUN = mean)
```

Doing the same with a `plyr` function:

```{r}
ddply(pDat, ~ gType, summarize, gExp = mean(gExp))
```

Let's make a stripplot so we can sanity test our $t$ test result.
And again with `ggplot2`:

```{r}
ggplot(pDat, aes(x = gExp, y = gType)) + geom_point()
```

We will do a two-sample $t$ test comparing wild type to the Nrl knockouts.

```{r}
t.test(gExp ~ gType, pDat)
```

If we save the $t$ test result, we can inspect what it is.

```{r}
ttRes <- t.test(gExp ~ gType, pDat)
str(ttRes)
```
It is a `list`; some of the more interesting and useful components are the test statistic and p-value. Recall how to extract components of a list:
```{r}
ttRes$statistic
ttRes$p.value
```


## Two sample tests -- a handful of genes

Let's grab the data from 6 genes. I've picked them for you: 3 are interesting ('hits'), 3 are not. 
```{r}
keepGenes <- c("1431708_a_at", "1424336_at", "1454696_at",
               "1416119_at", "1432141_x_at", "1429226_at" )
```



```{r}
prepareData <- function(jDat, design){
  jDat <- jDat[, design$sidChar]
  newDat <- cbind(design, t(jDat))
  
  # in case of multiple probes selected
  newDat <- melt(newDat, 
                 measure.vars=grep("_at", colnames(newDat), value=TRUE))
  newDat <- rename(newDat, c("value"="gExp",
                             "variable"="gene"))  # rename column
  return(newDat)
}

miniDat <- prepareData(subset(prDat, rownames(prDat) %in% keepGenes),
                       prDes)
str(miniDat)
```

Plot the genes:
```{r}
ggplot(miniDat, aes(x = gExp, y = gType, color = gType)) +
  facet_wrap(~ gene, scales="free_x") +
  geom_point(alpha = 0.7) +
  theme(panel.grid.major.x = element_blank())
```

Let's use data aggregation techniques to conduct some two group comparisons for each of these 6 genes. Recall the syntax of the two-sample t-test for one gene:

```{r eval=FALSE}
t.test(gExp ~ gType, someDat)
```

```{r}
d_ply(miniDat, ~ gene, function(x) t.test(gExp ~ gType, x), .print = TRUE)
d_ply(miniDat, ~ gene, function(x) wilcox.test(gExp ~ gType, x, exact=FALSE), .print = TRUE)
```

```{r}
d_ply(miniDat, ~ gene, function(x){
  ks.test(with(x, gExp[gType=="wt"]),
          with(x, gExp[gType=="NrlKO"]), exact=FALSE)
  }, .print = TRUE)
```

Use `dlply()` to retain everything in a new list with one component per probeset:
```{r}
ttRes <- dlply(miniDat, ~ gene, function(x) t.test(gExp ~ gType, x))
names(ttRes)
ttRes[["1454696_at"]]
```




## Ideas for take-home work

In our last example, can you edit the inner function to use the Wilcoxon or KS test? Or maybe do the $t$ test, the Wilcoxon, and the KS test and return all 3 p-values?

Conduct two group comparisons for all 6 genes, and store test statistics and the p-values via `ddply()`
```{r}
multiCompare <- function(jDat){
  ttRes <- ddply(jDat, ~ gene, function(z) {
    tz <- t.test(gExp ~ gType, z)
    wz <- wilcox.test(gExp ~ gType, z, exact=FALSE)
    kz <- suppressWarnings(
      ks.test(with(z, gExp[gType=="wt"]),
              with(z, gExp[gType=="NrlKO"]), exact=FALSE))
    
    c(tStat = tz$statistic, pVal.tStat = tz$p.value,
      wilcoxStat=wz$statistic, pVal.wilcoxStat = wz$p.value,
      ksStat=kz$statistic, pVal.ksStat = kz$p.value)
    })
  return(ttRes)
}
multiCompare(miniDat)
```

Scale up to more genes ... I'm thinking ~100. 


```{r}
set.seed(987)
keep <- sample(1:nrow(prDat), size=100)
head(keep)
miniDat2 <- prepareData(prDat[keep,], prDes)
```


Make a numeric matrix with p-values for many genes (= rows) for various two-group tests. Scatterplot them against each other. Are the p-values as similar as you'd expect? Do you need to log transform the axes to get more insight?

```{r}
tr2 <- multiCompare(miniDat2)
pvals2 <- tr2[,grep("pVal", colnames(tr2))]
splom(pvals2)
cor(pvals2)
```
The p-values don't seem very correlated between different tests.

Try log2 transform the p-values, because I don't care much whether those large p-values are correlated, instead I'm interested in whether small p-values that indicates significance is consistent between different tests. Log2 transform reduces the effect of large p-values.
```{r}
splom(log2(pvals2), panel=panel.smoothScatter)
cor(log2(pvals2))
```


Convert your numeric matrix of p-values into a matrix of `TRUE/FALSE` or zeros and ones by hard-threshholding, e.g. at the conventional 0.05 level. Use `apply()` or a function from `plyr` to make some interesting row or column summaries. How many genes are significant according to the different tests? For each gene, how many of the tests return a significant p-value? How many genes are "hits" by all 3 methods, by exactly 2, by exactly 1 and by none?

```{r}
sigCount <- apply(pvals2 <= 0.05, 1, sum)
sum(sigCount == 3)
sum(sigCount == 2)
sum(sigCount == 1)
```

Use `system.time()` to do a little timing study, doing differential expression analysis for gene sets of various sizes. I'm thinking 5, 10, 50, 100, 500 ... somthing like that. Don't immediately go all the way to 30K. We will use different strategies for our full scale-up. Plot the DE analysis time against the number of genes.

```{r}
for (n in c(5, 10, 50, 100, 500)){
  set.seed(987)
  keep <- sample(1:nrow(prDat), size=n)
  head(keep)
  miniDat2 <- prepareData(prDat[keep,], prDes)
  print(system.time(multiCompare(miniDat2)))
}

```

